Quantifying maternal reproductive output of chondrichthyan fishes

### 2024-05-31 rmds/manuscript.tex
revisions

For the live-bearing and egg-laying class of chondrichthyan fishes a three parameter logistic `maternity' function with a variable upper asymptote, \(P_{Max}\), can be used to predict the average probability of a female giving birth or laying eggs in a season. Although fundamental to calculating the reproductive capacity of a population, few studies report maternity functions, with maturity functions often used as a proxy. Applying logistic models to simulated and empirical data showed that it was feasible to estimate \(P_{Max}\) from maternal data and that accuracy, bias, and confidence interval coverage often improved compared to when a fixed value was used. Maturity parameters could be estimated with greater accuracy; however, substituting them for maternity parameters overestimated lifetime reproductive output. Greater use of maternity functions has the potential to improve calculation of reproductive output in quantitative populations models. In addition to improvements in parameter estimation, this method involves fewer assumptions and enables statistical inferences to be made on frequency of reproduction. This study suggests that for sample sizes \textgreater{} 100, estimation of \(P_{Max}\) is generally preferable to using a fixed value.

\begin{itemize}
\tightlist
\item
  A sentence with caveat about the need for large sample sizes to produce ``feasible'' results is warranted.
\end{itemize}

@@

The effect of the different methods in ultimately calculating \(R_0\) varied considerably across the various scenarios. For school sharks, estimating \(P_{Max}\), was the most effective way to minimize bias in most scenarios (Figure S10). Substituting maturity parameters (2PLF-maturity method) led to a median overestimation of \(R_0\) by 3.4\% across all scenarios, while the 2PLF-maternity method led to a median underestimate of 11.1\%. (Figure S10). In contrast, for the gummy shark substituting maturity parameters led to a median overestimation of \(R_0\) of just 1\% across all scenarios which outperformed the 3PLF methods in several instances (Figure S10). Using the 2PLF method for gummy sharks led to a median overestimate of \(R_0\) of 22\% (Figure S10). For both species, better accuracy in calculating \(R_0\) was achieved with the 3PLF methods, with the 3PLF-estimated method performing best in most scenarios tested (Figure 6, Figure S11). For gummy sharks the 3PLF-fixed method performed best at lower sample sizes and lower values of \(P_{Max}\).

@@

Relatively few reproductive biology studies have used maternity functions to model maternal reproductive output in chondrichthyan fishes. Where they have been employed the approach has typically been to use a three parameter logistic function with a fixed, user-defined value for the upper asymptote, \(P_{Max}\). Here we show that it is feasible to estimate \(P_{Max}\) from maternal data, in turn enabling statistical inferences on reproductive periodicity. Applying 3PLF models with estimated and fixed values of \(P_{Max}\) to simulated data showed that accuracy, bias, and confidence interval coverage often improved when \(P_{Max}\) was estimated. Using a fixed value for \(P_{Max}\) in some cases resulted in lower bias at low sample sizes. Overall, however, relatively large sample sizes were needed to accurately estimated \(P_{Max}\), particularly in species with longer reproductive cycles. This study also demonstrated that, in some cases, recruitment can be overestimated when maturity data were used to approximate maternal data and was inaccurate if \(P_{Max}\) was ignored when estimating maternity parameters. Based on these findings we outline considerations for practitioners using these methods and illustrate how they can provide novel insights into reproductive biology. We conclude by discussing the advantages of adopting this approach to quantifying maternal reproductive output and future directions.

@@

Using simulated data to compare the relative performance of the 3PLF-estimated and 3PLF-fixed methods subject to a range of variables showed that it was feasible to estimate \(P_{Max}\), but also identified situations where it may be preferable to fix \(P_{Max}\). While the estimation of maternity parameters in most simulations was often possible, it was difficult to do so accurately. Outcomes of the simulation study showed that, for non-annual reproductive cycles, \(\geq100\) and \(\geq200\) maternal females were typically needed to estimate \(P_{Max}\) with \textless10\% MARE for school and gummy sharks, respectively (Figure 5). This is a much larger error than is achievable from conventional logistic maturity analysis (Roa et al. 1999). In the case of the gummy shark this inaccuracy led to the 2PLF-maturity method performing comparably or better than the 3PLF methods in ultimately quantifying \(R_0\). Relatively wide confidence intervals were also obtained for \(P_{Max}\) in the empirical analysis for sandbar sharks, even with a total sample size of \textgreater{} 1000 females, of which 206 were in maternal condition. A similar sample size was used by Colonello \emph{et al}. (2016) to estimate plausible values of \(P_{Max}\) and maternity parameters for south Atlantic spiny dogfish.

Providing clear guidelines on when to fix or estimate \(P_{Max}\) is difficult as model performance was species-specific and influenced by life history traits and gear selectivity. Most instances where the 3PLF-fixed method outperformed the 3PLF-estimated method typically occurred at lower sample sizes. At sample sizes \textless100 it is therefore advisable to fix \(P_{Max}\). It is also important to note that accurate estimation of the \(P_{Max}\) parameter required a sample size of 100 - 200 females in maternal condition in simulated data, and that larger sample sizes are needed with longer reproductive cycles.

These results suggest that data requirements of the 3PLF analyses may be prohibitively large for many chondrichthyans and therefore best suited to use on commercially captured species where large sample sizes can be obtained (Oddone et al. 2010, Tribuzio and Kruse 2012). For smaller samples sizes, placing a prior or penalty on \(P_{Max}\) may also help overcome issues of parameter inaccuracy, enabling existing information about reproductive periodicity to be incorporated from other populations or closely-related species (Smart and Grammer 2021). Information on length at maturity, which is likely to be relatively easier to estimate, may also be informative in the development of priors, with \(L_{50}\) effectively providing a lower bound for \(L'_{50}\).

For the numerous data-poor chondrichthyan species, or where data on pregnant females are sparse or unattainable, the current practice of using maturity parameters as a proxy for maternity parameters will likely still be the only option. From this perspective, the outperformance of the 3PLF-methods by the 2PLF-methods for the gummy shark in several simulations is encouraging. However, the extent to which maturity parameters can provide a good approximation of maternity parameters may be species-specific, depending on how close \(L_{50}\) is to \(L'_{50}\). While several studies have shown these parameters to be similar (Walker 2007, Soto-López et al. 2018), \(L'_{50}\) is more frequently shifted to the right of \(L_{50}\) (Braccini et al. 2006, Montealegre-Quijano et al. 2014, Colonello et al. 2016, Palacios-Hernández et al. 2020), and this length difference can equate to one or more years. For example, Harry et al. (2013) found both spot-tail shark, \emph{C. sorrah}, and Australian blacktip shark, \emph{C. tilstoni}, began reproducing the year after reaching sexual maturity. Similarly, Fujinami et al. (2017) found \(A'_{50}\) to be 1.4 years older than \(A_{50}\) in blue sharks.

@@

This unexpected result may be related to the gradual attainment of asymptotic maternal status. Maturity in sandbar sharks appears to occur over an extended size range. \(L_{5}\) and \(L_{95}\) are 146cm and 176cm -- a length interval that corresponds to around seven years given the sandbar shark's slow growth rate (Hale and Baremore 2013). With females maturing over a broad range of sizes and ages, it is possible that smaller females reproduce less frequently. The outcomes of this study suggest that larger sharks may ultimately be capable of reproducing biennially, although few females appear to attain this size (Hale and Baremore 2013). As such, the modal reproductive frequency is still at least triennial. Further work would be required to confirm this hypothesis, however, such an outcome would have important implications for management; not only are larger females more fecund (Baremore and Hale 2012), they also reproduce more frequently. This implies that size selective fishing or the removal of larger and older female sharks could have a disproportionately greater impact on population productivity.

@@

One benefit of this approach is that it reduces the need for subjective modelling assumptions relating to \(P_{Max}\). Such assumptions can have important implications for population modelling. For example, due to uncertainty in temporal and spatial frequency of reproduction in sparsely-spotted stingarees, \emph{Urolophus paucimaculatus}, Trinnie \emph{et al}. (Trinnie et al. 2014) estimated maternity parameters for ten plausible scenarios. In stock assessment of western North Atlantic sandbar sharks a breeding frequency of 2.5 years was assumed to account for uncertainty in the duration of the reproductive cycle (SEDAR 2017).

@@

>>>>

For the live-bearing and egg-laying class of chondrichthyan fishes a three parameter logistic `maternity' function with a variable upper asymptote, \(P_{Max}\), can be used to predict the average probability of a female giving birth or laying eggs in a season. Although fundamental to calculating the reproductive capacity of a population, few studies report maternity functions, with maturity functions often used as a proxy. Applying logistic models to simulated and empirical data showed that it was feasible to estimate \(P_{Max}\) from maternal data and that accuracy, bias, and confidence interval coverage often improved compared to when a fixed value was used. However, sample sizes of 100 - 200 maternal females were required for accurate estimation of \(P_{Max}\). Maturity parameters could be estimated with greater accuracy; however, substituting them for maternity parameters overestimated lifetime reproductive output. Greater use of maternity functions has the potential to improve calculation of reproductive output in quantitative populations models. In addition to improvements in parameter estimation, this method involves fewer assumptions and enables statistical inferences to be made on frequency of reproduction.

@@

The effect of the different methods in ultimately calculating \(R_0\) varied considerably across the various scenarios. For school sharks, estimating \(P_{Max}\), was the most effective way to minimize bias in most scenarios (Figure S10). Substituting maturity parameters (2PLF-maturity method) led to a median overestimation of \(R_0\) by 3.4\% across all scenarios, while the 2PLF-maternity method led to a median underestimate of 11.1\%. (Figure S10). In contrast, for the gummy shark substituting maturity parameters led to a median overestimation of \(R_0\) of just 1\% across all scenarios which outperformed the 3PLF methods in several instances (Figure S11). Using the 2PLF method for gummy sharks led to a median overestimate of \(R_0\) of 22\% (Figure S10). For both species, better accuracy in calculating \(R_0\) was achieved with the 3PLF methods, with the 3PLF-estimated method performing best in most scenarios tested (Figure 6, Figure S12). For gummy sharks the 3PLF-fixed method performed best at lower sample sizes and lower values of \(P_{Max}\).

@@

Relatively few reproductive biology studies have used maternity functions to model maternal reproductive output in chondrichthyan fishes. Where they have been employed the approach has typically been to use a three parameter logistic function with a fixed, user-defined value for the upper asymptote, \(P_{Max}\). Here we show that it is feasible to estimate \(P_{Max}\) from maternal data, in turn enabling statistical inferences on reproductive periodicity. Applying 3PLF models with estimated and fixed values of \(P_{Max}\) to simulated data showed that accuracy, bias, and confidence interval coverage often improved when \(P_{Max}\) was estimated. Using a fixed value for \(P_{Max}\) in some cases reduced bias. Overall, however, relatively large sample sizes were needed to accurately estimate \(P_{Max}\), particularly with longer reproductive cycles. This study also demonstrated that, in some cases, recruitment can be overestimated when maturity data were used to approximate maternal data and inaccurate if \(P_{Max}\) was ignored when estimating maternity parameters. Based on these findings we outline considerations for practitioners using these methods and illustrate how they can provide novel insights into reproductive biology. We conclude by discussing the advantages of adopting this approach and future directions.

@@

Using simulated data to compare the relative performance of the 3PLF-estimated and 3PLF-fixed methods subject to a range of variables showed that it was feasible to estimate \(P_{Max}\), but was difficult to do so accurately. For non-annual reproductive cycles, \(> 100\) school sharks and \(> 200\) gummy sharks in maternal condition were typically needed to estimate \(P_{Max}\) with \textless10\% MARE (Figure 5). This indicates that the both sample size requirements and levels of uncertainty are likely to be larger than conventional logistic maturity analysis (Roa et al. 1999). In the case of the gummy shark this inaccuracy led to the 2PLF-maturity method performing comparably or better than the 3PLF methods in minimizing bias in \(R_0\) (Figure S11) Relatively wide confidence intervals were also obtained for \(P_{Max}\) in the empirical analysis for sandbar sharks, even with a total sample size of \textgreater{} 1000 females, of which 206 were in maternal condition. A similar sample size was used by Colonello \emph{et al}. (2016) to estimate plausible values of \(P_{Max}\) and maternity parameters for south Atlantic spiny dogfish.

The simulation study also also identified some situations where it may be preferable to fix \(P_{Max}\), particularly at lower sample sizes (\(\le\) 100 individuals total). Providing clear guidelines on this is difficult as model performance was species-specific and influenced by life history and gear selectivity. The success of any analysis using a fixed value also ultimately depends on the analyst choosing a suitable value. The potential to introduce a greater level of bias through the incorrect selection of \(P_{Max}\) should also be considered if using this approach (Natanson et al. 2019).

Together, these results suggest that data requirements of the 3PLF analyses may be prohibitively large for many chondrichthyans and therefore best suited to use on commercially captured species where large sample sizes can be obtained (Oddone et al. 2010, Tribuzio and Kruse 2012). For smaller samples sizes, placing a prior or penalty on \(P_{Max}\) may also help overcome issues of parameter inaccuracy, enabling existing information about reproductive periodicity to be incorporated from other populations or closely-related species (Smart and Grammer 2021). Information on length at maturity, which is likely to be relatively easier to estimate, may also be informative in the development of priors, with \(L_{50}\) effectively providing a lower bound for \(L'_{50}\).

For data-poor chondrichthyan species, or where data on pregnant females are sparse or unattainable, the current practice of using maturity parameters as a proxy for maternity parameters will likely still be the only option. From this perspective, the outperformance of the 3PLF-methods by the 2PLF-methods for the gummy shark in several simulations is encouraging. However, the extent to which maturity parameters can provide a good approximation of maternity parameters may be species-specific, depending on how close \(L_{50}\) is to \(L'_{50}\). While several studies have shown these parameters to be similar (Walker 2007, Soto-López et al. 2018), \(L'_{50}\) is more frequently shifted to the right of \(L_{50}\) (Braccini et al. 2006, Montealegre-Quijano et al. 2014, Colonello et al. 2016, Palacios-Hernández et al. 2020), and this length difference can equate to one or more years. For example, Harry et al. (2013) found both spot-tail shark, \emph{C. sorrah}, and Australian blacktip shark, \emph{C. tilstoni}, began reproducing the year after reaching sexual maturity. Similarly, Fujinami et al. (2017) found \(A'_{50}\) to be 1.4 years older than \(A_{50}\) in blue sharks.

@@

This unexpected result may be related to the gradual attainment of asymptotic maternal status. Maturity in sandbar sharks appears to occur over an extended size range. \(L_{5}\) and \(L_{95}\) are 146cm and 176cm -- a length interval that corresponds to around seven years given the sandbar shark's slow growth rate (Hale and Baremore 2013). With females maturing over a broad range of sizes and ages, it is possible that smaller females reproduce less frequently. The outcomes of this study suggest that larger sharks may ultimately be capable of reproducing biennially. Given that few females appear to attain this size (Hale and Baremore 2013). this suggests the modal reproductive frequency likely still at least triennial. Further work would be required to confirm this hypothesis, however, such an outcome would have important implications for management; not only are larger females more fecund (Baremore and Hale 2012), they also reproduce more frequently. This implies that size selective fishing or the removal of larger and older female sharks could have a disproportionately greater impact on population productivity.

@@

One benefit of this approach is that it reduces the need for subjective modelling assumptions relating to \(P_{Max}\). Such assumptions can have important implications for population modelling. For example, due to uncertainty in temporal and spatial frequency of reproduction in sparsely-spotted stingarees, \emph{Urolophus paucimaculatus}, Trinnie \emph{et al}. (2014) estimated maternity parameters for ten plausible scenarios. In stock assessment of western North Atlantic sandbar sharks a breeding frequency of 2.5 years was assumed to account for uncertainty in the duration of the reproductive cycle (SEDAR 2017).

@@

\bibitem[\citeproctext]{ref-natanson_presence_2019}
Natanson, L.J., Deacy, B.M., Joyce, W., and Sulikowski, J. 2019. Presence of a resting population of female porbeagles ({Lamna} nasus), indicating a biennial reproductive cycle, in the western {North} {Atlantic} {Ocean}. Fish. B-noaa. \textbf{117}.
